\documentclass{article}
\usepackage{neb-titles}
\usepackage{flexfig}

\begin{document}

\ActivityTitle[class=Statistics, number=1, name={Basic Parameters (Solutions)}]

\begin{enumerate}
\item Find the mean of the following list of numbers.

\begin{center}
6, 11, 9, 13, 10, 14, 15, 9
\end{center}

\textbf{Solution:} Remember that to find the mean of $n$ numbers $x_1$, $x_2$, $\ldots$, $x_n$, we add them up and divide by the number of numbers. In this case,

\begin{eqnarray*}
\frac{6 + 11 + 9 + 13 + 10 + 14 + 15 + 9}{8} & = & \frac{87}{8} \\
 & \approx & 10.8.
\end{eqnarray*}


  
\vspace{1cm}

\item Find the mean of the following list of numbers.

\begin{center}
75, 95, 62, 53, 92, 82, 52, 88, 79, 59
\end{center}

\textbf{Solution:} Remember that to find the mean of $n$ numbers $x_1$, $x_2$, $\ldots$, $x_n$, we add them up and divide by the number of numbers. In this case,

\begin{eqnarray*}
\frac{75 + 95 + 62 + 53 + 92 + 82 + 52 + 88 + 79 + 59}{10} & = & \frac{737}{10} \\
 & \approx & 73.7.
\end{eqnarray*}


  
\vspace{1cm}

\item Find the mean of the following list of numbers.

\begin{center}
1, 4, 1, 3, 0, 1, 4, 4
\end{center}

\textbf{Solution:} Remember that to find the mean of $n$ numbers $x_1$, $x_2$, $\ldots$, $x_n$, we add them up and divide by the number of numbers. In this case,

\begin{eqnarray*}
\frac{1 + 4 + 1 + 3 + 0 + 1 + 4 + 4}{8} & = & \frac{18}{8} \\
 & \approx & 2.2.
\end{eqnarray*}


  
\vspace{1cm}

\item Find the mean deviation of the following list of numbers.

\begin{center}
10, 9, 8, 5, 15
\end{center}

\textbf{Solution:} Remember that the mean deviation of $x_1$, $x_2$, $\ldots$, $x_n$ is \[ \frac{1}{n} \sum_{i=1}^n |x_i - \overline{x}|, \] where $\overline{x}$ is the mean of the $x_i$. In this case the mean is $\overline{x} = 9.40$. Then the mean deviation is

\begin{eqnarray*}
 & & \frac{1}{5} \big( |10 - 9.40| + |9 - 9.40| + |8 - 9.40| + |5 - 9.40| + |15 - 9.40| \big) \\
 & = & \frac{1}{5} \big( 0.60 + 0.40 + 1.40 + 4.40 + 5.60 \big) \\
 & = & 2.48
\end{eqnarray*}


  
\vspace{1cm}

\item Find the mean deviation of the following list of numbers.

\begin{center}
5, 2, 4, 5, 8, 6
\end{center}

\textbf{Solution:} Remember that the mean deviation of $x_1$, $x_2$, $\ldots$, $x_n$ is \[ \frac{1}{n} \sum_{i=1}^n |x_i - \overline{x}|, \] where $\overline{x}$ is the mean of the $x_i$. In this case the mean is $\overline{x} = 5.00$. Then the mean deviation is

\begin{eqnarray*}
 & & \frac{1}{6}\big( |5 - 5.00| + |2 - 5.00| + |4 - 5.00| + |5 - 5.00| + |8 - 5.00| + |6 - 5.00| \big) \\
 & = & \frac{1}{6}\big( 0.00 + 3.00 + 1.00 + 0.00 + 3.00 + 1.00 \big) \\
 & = & 1.33
\end{eqnarray*}


  
\vspace{1cm}

\item Find the standard deviation of the following list of numbers.

\begin{center}
14, 7, 13, 9, 6
\end{center}

\textbf{Solution:} Remember that the standard deviation of $x_1$, $x_2$, $\ldots$, $x_n$ is \[ \sqrt{\frac{\sum_{i=1}^n (x_i - \overline{x})^2}{n-1}}, \] where $\overline{x}$ is the mean of the $x_i$. In this case the mean is $\overline{x} = 9.80$. Then the standard deviation is

\begin{eqnarray*}
 & & \sqrt{\frac{1}{4}\big( (14 - 9.80)^2 + (7 - 9.80)^2 + (13 - 9.80)^2 + (9 - 9.80)^2 + (6 - 9.80)^2 \big)} \\
 & = & \sqrt{\frac{1}{4}\big( 17.64 + 7.84 + 10.24 + 0.64 + 14.44 \big)} \\
 & = & 3.18
\end{eqnarray*}


  
\vspace{1cm}

\item Suppose we have collected the following list of numbers.

\begin{center}
7, 5, 16, 8, 14, 10, 24, 6, 20, 20
\end{center}

Compute the z-scores of 3 and 24 with respect to this list.

\textbf{Solution:} Remember that the z-score of a particular number $x$ with respect to a list of numbers is \[ z = \frac{x - \overline{x}}{s}, \] where $\overline{x}$ is the mean and $s$ the standard deviation. In this case we can see that $\overline{x} = 13.0$ and $s = 6.418$, so that the z-score of 3 is \[ \frac{3 - 13.0}{6.418} = -1.557 \] and of 24 is  \[ \frac{24 - 13.0}{6.418} = 1.713. \]


  
\vspace{1cm}

\item Suppose we have collected the following list of numbers.

\begin{center}
8, 4, 8, 7, 8, 5, 9, 6, 9, 8
\end{center}

Compute the z-scores of 2 and 14 with respect to this list.

\textbf{Solution:} Remember that the z-score of a particular number $x$ with respect to a list of numbers is \[ z = \frac{x - \overline{x}}{s}, \] where $\overline{x}$ is the mean and $s$ the standard deviation.
In this case we can see that $\overline{x} = 7.2$ and $s = 1.600$, so that the z-score of 2 is
 \[ \frac{2 - 7.2}{1.600} = -3.249 \]
 and of 14 is  \[ \frac{14 - 7.2}{1.600} = 4.249. \]


  
\vspace{1cm}

\item Find the coefficient of variation of the following list of numbers.

\begin{center}
12, 11, 15, 19, 19, 20
\end{center}

\textbf{Solution:} Remember that the coefficient of variation of a list of numbers is $100\% \cdot s/\overline{x}$, where $s$ is the standard deviation and $\overline{x}$ the mean, expressed as a percentage. In this case the mean is $\overline{x} = 16.0$ and the standard deviation is $s = 3.55$, so the coefficient of variation is $100\% \cdot s/\overline{x} = 22\%$.


  
\vspace{1cm}

\item Find the coefficient of variation of the following list of numbers.

\begin{center}
8, 5, 3, 10, 9, 5, 5
\end{center}

\textbf{Solution:} Remember that the coefficient of variation of a list of numbers is $100\% \cdot s/\overline{x}$, where $s$ is the standard deviation and $\overline{x}$ the mean, expressed as a percentage. In this case the mean is $\overline{x} = 6.4$ and the standard deviation is $s = 2.38$, so the coefficient of variation is $100\% \cdot s/\overline{x} = 37\%$.


  
\vspace{1cm}
\end{enumerate}

\end{document}
